Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-16 manuscript.qmd
plot joint state dist as algebra equation
Related: #11

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, which is illustrated in @fig-model-state-dist. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx. @fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced??? T-category tumors. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig_model_state_dist.svg){#fig-model-state-dist}

>>>>

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx. @fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced??? T-category tumors. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention. The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively. In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension. In the center, the time-prior is plotted as a diagonal matrix. The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}
